---
title: "Capstone project"
author: "Rebecca Laes-Kushner"
format: html
editor: visual
source: charity:water
---

## Description

This project analyses baseline data from a water, sanitation and hygiene (WASH) project in Bangladesh run by charity:water. The survey asks about access to water, water issues, handwashing practices and defecation (pooping) practices. The long-term goal would be to compare baseline with endline data to see changes in each town due to the WASH project, such as building public latrines or boreholes and providing hygiene education.

## Libraries

```{r}
#| message: false
# Install these once in the Console (not here), then just load them:
# install.packages(c("tidyverse", "here", "readxl", "gt"))
# Note: dplyr and readr come with tidyverse, so loading tidyverse is enough.
library(here)
library(tidyverse)
library(readxl)
library(gt)
```

## Data import

```{r}
raw <- read_excel(
  here::here("data/raw/Final Baseline and_HwiseSatisfaction to upload raw.xlsx"),
  na = c("", "NA", "N/A", "."))

# Replace NA with 0 in every numeric column. `across(where(is.numeric), ...)`
# picks the numeric columns; `replace_na(.x, 0)` swaps the NAs for 0. This is
# the line you already had, and it does work, so we keep just this one.
raw2 <- raw |>
  mutate(across(where(is.numeric), ~ replace_na(.x, 0)))

# A caution worth keeping in mind: turning NA into 0 is a real decision, not
# just a fix. 0 means "this value is zero", which is different from "we don't
# know". For an average like avg_time_get_water below, leaving the NAs in and
# using mean(..., na.rm = TRUE) is usually more honest than first making them 0,
# because a 0 would pull the average down. So apply this only to columns where 0
# is genuinely the right value (for example a count of containers that were
# simply not owned).

str(raw2)
```

## Using pre-prepared Excel data

So in order to move forward, I replaced nulls with 0 in Excel and imported "na" version (or "no na" version).

```{r}
rawna <- read_excel(
  here::here("data/raw/Final Baseline and_HwiseSatisfaction to upload no NA.xlsx")
)
```

## Prepare data

```{r}
# Build the per-household (per-row) variables.
#
# avg_time_get_water: you want the average time PER TOWN, so we group_by(upazila)
# first; then mean() inside mutate() gives each row its town's average. round(x, 1)
# gives one decimal (your earlier round(...),3) put the 3 outside round(), so it
# never reached the function). ungroup() at the end clears the grouping so the
# later steps are not still grouped by accident.
processedna <- rawna |>
  group_by(upazila) |>
  mutate(avg_time_get_water = round(mean(total_time_to_get_water, na.rm = TRUE), 1)) |>
  ungroup() |>
  relocate(avg_time_get_water, .after = total_time_to_get_water) |>
  # total water per household (litres), then per capita
  mutate(total_water = (num_25L_filled * 25) + (num_20L_filled * 20) +
                       (num_15L_filled * 15) + (num_10L_filled * 10) +
                       (num_5L_filled  *  5)) |>
  relocate(total_water, .after = other_containers) |>
  mutate(water_per_cap = total_water / num_hh_members) |>
  relocate(water_per_cap, .after = total_water)
```

Now summarise by town. The key fix here is the pipe (`|>`) after every line of
the chain except the last; a missing pipe is what produced the
`argument ".data" is missing` error before.

```{r}
avg_water_time_upazila <- processedna |>
  group_by(upazila) |>
  summarize(
    mean_time          = mean(avg_time_get_water, na.rm = TRUE),
    mean_water         = mean(total_water,        na.rm = TRUE),
    mean_water_per_cap = mean(water_per_cap,      na.rm = TRUE)
  )

avg_water_time_upazila
```

To inspect a single column, reach into the data frame with `$` (writing
`str(upazila)` on its own fails because `upazila` is a column, not a standalone
object):

```{r}
str(processedna$upazila)
```

A formatted summary table with `gt`:

```{r}
avg_water_time_upazila |>
  gt() |>
  fmt_number(
    columns  = c(mean_time, mean_water, mean_water_per_cap),
    decimals = 1
  )
```

## 
